### 2014-05-14 doc/thesis/review.tex
Trying to improve a little bit the text about gradual typing

\emph{Pluggable type systems} \citep{bracha2004pluggable} generalize
the idea of optional type systems that Strongtalk put in practice,
to have different optional type systems that can be layered on top
of a dynamically typed language without affecting its original run-time
semantics.
Although these systems can be unsound on their interaction with the
dynamically typed part of the language or even by design, their
unsoundness does not affect run-time safety, as the language run-time
semantics still cathes any run-time errors caused by an unsound
type system.

@@

The main goal of gradual typing \citep{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, \citet{siek2006gradual} extended the simply typed
lambda calculus with the dynamic type $?$.
In gradual typing, type annotations are optional, so an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

\begin{figure}[!ht]
\textbf{Syntax of the gradually-typed lambda calculus}\\
\dstart
$$
\begin{array}{rlr}
T ::= & & \emph{types}\\
& \;\; \Number & \emph{base type number}\\
& | \; \String & \emph{base type string}\\
& | \; ? & \emph{dynamic type}\\
& | \; T \rightarrow T & \emph{function types}\\
e ::= & & \emph{expressions}\\
& \;\; c & \emph{constants}\\
& | \; x & \emph{variables}\\
& | \; \lambda x{:}T{.}e & \emph{abstractions}\\
& | \; e_{1} e_{2} & \emph{application}
\end{array}
$$
\dend
\caption{Syntax of the gradually-typed lambda calculus}
\label{fig:gtlc}
\end{figure}

The central idea of gradual typing is the \emph{consistency}
relation, written $\tau_{1} \sim \tau_{2}$, that allows implicit
conversions to and from the dynamic type, and disallows conversions
between inconsistent types \citep{siek2006gradual}.
For instance, $\mathbf{int} \sim \;?$, $? \sim \mathbf{int}$,
$\mathbf{string} \sim \;?$, and $? \sim \mathbf{string}$,
but $\mathbf{int} \not\sim \mathbf{string}$, and
$\mathbf{string} \not\sim \mathbf{int}$.
The consistency relation is neither commutative nor transitive.

\begin{figure}[!ht]
\textbf{Type consistency}\\
\dstart
$$
\begin{matrix}
\Number \sim \Number \mylabel{C-NUM} \;\;\; \String \sim \String \mylabel{C-STR}
\\ \\
T \sim \;? \mylabel{C-ANY1} \;\;\; ? \sim T \mylabel{C-ANY2}
\\ \\
\dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
\end{matrix}
$$
\dend
\caption{The consistency relation}
\label{fig:consistency}
\end{figure}

Gradual typing is similar to the approaches proposed by
\citet{abadi1989dts} and \citet{thatte1990qst} by including a
dynamic type to a statically typed language.
However, these three approaches differ in the way they handle the
dynamic type.
While \citet{siek2006gradual} rely on the consistency relation,
\citet{abadi1989dts} rely on type equality with explicit projections
and injections, and \citet{thatte1990qst} relies on subtyping.
The subtyping relation actually is a pitfall on Thatte's quasi-static
typing, because the dynamic type is put as the top and the bottom of
the subtyping relation.
Therefore, downcasts combined with the transitivity of subtyping
accepts programs that should be rejected.

\begin{figure}[!ht]
\textbf{Gradual Type System}\\
\dstart
$$
\begin{matrix}
\dfrac{\env(x) = T}
      {\env \vdash x:T} \mylabel{T-VAR}
\;\;\;
\dfrac{type(c) = T}
      {\env \vdash c:T} \mylabel{T-CONST}
\\ \\
\dfrac{\env, x:T_{1} \vdash e:T_{2}}
      {\env \vdash \lambda x:T_{1}.e:T_{1} \rightarrow T_{2}} \mylabel{T-ABS}
\;\;\;
\dfrac{\env \vdash e_{1}:\;? \;\;\;
       \env \vdash e_{2}:T}
      {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
\\ \\
\dfrac{\env \vdash e_{1}:T_{1} \rightarrow T_{2} \;\;\;
       \env \vdash e_{2}:T_{3} \;\;\;
       T_{3} \sim T_{1}}
      {\env \vdash e_{1} e_{2}:T_{2}} \mylabel{T-APP2}
\end{matrix}
$$
\dend
\caption{Typing rules of gradually-typed lambda calculus}
\label{fig:gts}
\end{figure}

Later, \citet{siek2007objects} discovered that the consistency relation
is orthogonal to subtyping, so we can combine the two relations to
achieve both consistency and subtyping.
The \emph{consistent-subtyping} relation,
written $\tau_{1} \lesssim \tau_{2}$,
is essential to design object-oriented languages that have a
gradual type system.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.

\begin{figure}[!ht]
\textbf{Consistent-Subtyping}\\
\dstart
$$
\begin{matrix}
\Number \lesssim \Number \mylabel{C-NUM} \;\;\; \String \lesssim \String \mylabel{C-STR}
\\ \\
T \lesssim \;? \mylabel{C-ANY1} \;\;\; ? \lesssim T \mylabel{C-ANY2}
\\ \\
\dfrac{T_{3} \lesssim T_{1} \;\;\; T_{2} \lesssim T_{4}}
      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
\;\;\;
\dfrac{\forall i \in 1..n \;\;\; T_{i} \lesssim T_{i}'}
      {[l_{i}:T_{i}^{\;i\in1..n+m}] \lesssim [l_{i}:T_{i}'^{\;i\in1..n}]} \mylabel{C-OBJ}
\end{matrix}
$$
\dend
\caption{The consistent-subtyping relation}
\label{fig:consistent-subtyping}
\end{figure}

Another important feature of gradual typing is the theoretic
foundation that it provides for inserting run-time checks that
prove dynamically typed code does not violate the invariants of
statically typed code, thus preserving type safety.
To do that, \citet{siek2006gradual} defined the run-time semantics
of gradual typing as a translation to an intermediate language with
explicit casts at the frontiers between statically and dynamically
typed code.
The semantics of these casts is based on the higher-order contracts
proposed by \citet{findler2002chf}.

@@

Still, we can view gradual typing as a way to formalize an optional
type system when the gradual type system does not insert run-time
checks.
BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
are two examples of object-oriented languages that use gradual
typing to formalize their optional type systems.
The optional type systems of both BabyJ and Alore are nominal.
BabyJ uses the relation $\approx$ that is similar to the consistency
relation while Alore combines  subtyping along with the consistency
relation to define a \emph{consistent-or-subtype} relation.
The consistent-or-subtype relation is different from the
consistent-subtyping relation of \citet{siek2007objects}, but it is
also written $\tau_{1} \lesssim \tau_{2}$.
The consistent-or-subtype relation holds when $\tau_{1} \sim \tau_{2}$
or $\tau_{1} <: \tau_{2}$, where $<:$ is transitive and $\sim$ is not.
Alore also extends its optional type system to include optional
monitoring of run-time type errors in the gradual typing style.

>>>>

\emph{Pluggable type systems} \citep{bracha2004pluggable} generalize
the idea of optional type systems that Strongtalk put in practice,
to have different optional type systems that can be layered on top
of a dynamically typed language without affecting its original run-time
semantics.
Although these systems can be unsound on their interaction with the
dynamically typed part of the language or even by design, their
unsoundness does not affect run-time safety, as the language run-time
semantics still catches any run-time errors caused by an unsound
type system.

@@

The main goal of gradual typing \citep{siek2006gradual} is to allow
programmers to choose between static and dynamic typing in the same
language.
To do that, \citet{siek2006gradual} extended the simply typed
lambda calculus with the dynamic type $?$, as we shall see in
Figure \ref{fig:gtlc}.
In gradual typing, type annotations are optional, so an untyped
variable is syntactic sugar for a variable whose declared type is
the dynamic type $?$, that is, $\lambda x.e \equiv \lambda x{:}?.e$.
Under these circumstances, we view gradual typing as a way to add
a dynamic type to statically typed languages.

\begin{figure}[!ht]
\dstart
$$
\begin{array}{rlr}
T ::= & & \emph{types}\\
& \;\; \Number & \emph{base type number}\\
& | \; \String & \emph{base type string}\\
& | \; ? & \emph{dynamic type}\\
& | \; T \rightarrow T & \emph{function types}\\
e ::= & & \emph{expressions}\\
& \;\; c & \emph{constants}\\
& | \; x & \emph{variables}\\
& | \; \lambda x{:}T{.}e & \emph{abstractions}\\
& | \; e_{1} e_{2} & \emph{application}
\end{array}
$$
\dend
\caption{Syntax of the gradually-typed lambda calculus}
\label{fig:gtlc}
\end{figure}

The central idea of gradual typing is the \emph{consistency}
relation, written $T_{1} \sim T_{2}$.
The consistency relation allows implicit conversions to and from the
dynamic type, and disallows conversions between inconsistent types
\citep{siek2006gradual}.
For instance, $\Number \sim \;?$, $? \sim \Number$,
$\String \sim \;?$, and $? \sim \String$,
but $\Number \not\sim \String$, and
$\String \not\sim \Number$.
The consistency relation is both reflexive and symmetric, but
it is neither commutative nor transitive.

\begin{figure}[!ht]
\dstart
$$
\begin{matrix}
\Number \sim \Number \mylabel{C-NUM} \;\;\; \String \sim \String \mylabel{C-STR}
\\ \\
T \sim \;? \mylabel{C-ANY1} \;\;\; ? \sim T \mylabel{C-ANY2}
\\ \\
\dfrac{T_{3} \sim T_{1} \;\;\; T_{2} \sim T_{4}}
      {T_{1} \rightarrow T_{2} \sim T_{3} \rightarrow T_{4}} \mylabel{C-FUN}
\end{matrix}
$$
\dend
\caption{The consistency relation}
\label{fig:consistency}
\end{figure}

Figure \ref{fig:consistency} defines the consistency relation.
Rules \textsc{C-NUM} and \textsc{C-STR} are reflexive. 
Rules \textsc{C-ANY1} and \textsc{C-ANY2} are the rules that allow
conversions to and from the dynamic type $?$.
The rule \textsc{C-FUN} resembles the subtyping of functions
because it is contravariant on the argument type and covariant on the
return type.

Figure \ref{fig:gts} uses the consistency relation in the typing rules
of the gradual type system of the simply typed lambda calculus extended
with the dynamic type $?$.
The environment $\env$ is a function from variables to types, and
the directive $type$ is a function from constants to types.
The rule \textsc{T-VAR} uses the environment function $\env$ to get the
type of a variable $x$.
The rule \textsc{T-CONST} uses the directive $type$ to get the type of
a constant $c$.
The rule \textsc{T-ABS} evaluates the expression $e$ with an environment
$\env$ that binds the variable $x$ to the type $T_{1}$, and the resulting
type is the the function type $T_{1} \rightarrow T_{2}$.
The rule \textsc{T-APP1} handles function calls where the type of the
function is dynamically typed; in this case, the argument type may have
any type and the resulting type has the dynamic type.
The rule \textsc{T-APP2} handles function calls where the type of the
function is statically typed; in this case, the argument type should
be consistent to the argument type of the function's signature.

\begin{figure}[!ht]
\dstart
$$
\begin{matrix}
\dfrac{\env(x) = T}
      {\env \vdash x:T} \mylabel{T-VAR}
\;\;\;
\dfrac{type(c) = T}
      {\env \vdash c:T} \mylabel{T-CONST}
\\ \\
\dfrac{\env, x:T_{1} \vdash e:T_{2}}
      {\env \vdash \lambda x:T_{1}.e:T_{1} \rightarrow T_{2}} \mylabel{T-ABS}
\;\;\;
\dfrac{\env \vdash e_{1}:\;? \;\;\;
       \env \vdash e_{2}:T}
      {\env \vdash e_{1} e_{2}:\;?} \mylabel{T-APP1}
\\ \\
\dfrac{\env \vdash e_{1}:T_{1} \rightarrow T_{2} \;\;\;
       \env \vdash e_{2}:T_{3} \;\;\;
       T_{3} \sim T_{1}}
      {\env \vdash e_{1} e_{2}:T_{2}} \mylabel{T-APP2}
\end{matrix}
$$
\dend
\caption{Gradual type system gradually-typed lambda calculus}
\label{fig:gts}
\end{figure}

Gradual typing is similar to the approaches proposed by
\citet{abadi1989dts} and \citet{thatte1990qst} by including a
dynamic type to a statically typed language.
However, these three approaches differ in the way they handle the
dynamic type.
While \citet{siek2006gradual} rely on the consistency relation,
\citet{abadi1989dts} rely on type equality with explicit projections
and injections, and \citet{thatte1990qst} relies on subtyping.

The subtyping relation $\subtype$ actually is a pitfall on Thatte's
quasi-static typing, because quasi-static typing sets the dynamic type
as the top and the bottom of the subtying relation:
$T \subtype \;?$ and $? \subtype T$.
Subtyping is transitive, so we have that
\[
\frac{\Number \subtype \;? \;\;\;
      ? \subtype \String}
     {\Number \subtype \String}
\]
Therefore, downcasts combined with the transitivity of subtyping
accepts programs that should be rejected.

Later, \citet{siek2007objects} reported that the consistency relation
is orthogonal to subtyping, so we can combine the two relations to
achieve both consistency and subtyping.
The \emph{consistent-subtyping} relation,
written $T_{1} \lesssim T_{2}$,
is the foundation to design object-oriented languages that have a
gradual type system.
Like the consistency relation, and unlike the subtyping relation,
the consistent-subtyping relation is not transitive.
Therefore, $\Number \lesssim \;?$, $? \lesssim \Number$,
$\String \lesssim \;?$, and $? \lesssim \String$,
but $\Number \not\lesssim \String$, and
$\String \not\lesssim \Number$.

So far, gradual typing looks like a mere formalization to optional
type systems, as a gradual type system uses the consistency or
consistent-subtyping relation to statically check the interaction
between statically and dynamically typed code, and not influencing
the run-time semantics.

However, another important feature of gradual typing is the theoretic
foundation that is provides for inserting run-time checks that
prove dynamically typed code does not violate the invariants of
statically typed code, thus preserving type safety.
To do that, \citet{siek2006gradual,siek2007objects} defined the
run-time semantics of gradual typing as a translation to an
intermediate language with explicit casts at the frontiers between
statically and dynamically typed code.
The semantics of these casts is based on the higher-order contracts
proposed by \citet{findler2002chf}.

@@

Still, we can view gradual typing as a way to formalize an optional
type system when the gradual type system does not insert run-time
checks.
BabyJ \citep{anderson2003babyj} and Alore \citep{lehtosalo2011alore}
are two examples of object-oriented languages that use gradual
typing to formalize their optional type systems.
The optional type systems of both BabyJ and Alore are nominal.
BabyJ uses the relation $\approx$ that is similar to the consistency
relation while Alore combines  subtyping along with the consistency
relation to define a \emph{consistent-or-subtype} relation.
The consistent-or-subtype relation is different from the
consistent-subtyping relation of \citet{siek2007objects}, but it is
also written $T_{1} \lesssim T_{2}$.
The consistent-or-subtype relation holds when $T_{1} \sim T_{2}$
or $T_{1} <: T_{2}$, where $<:$ is transitive and $\sim$ is not.
Alore also extends its optional type system to include optional
monitoring of run-time type errors in the gradual typing style.
